Mapping Cancer: the Potential of Cartograms and Alternative Map Displays

### 2020-05-15 literature.Rmd
Edited according to reviewers suggestions up to Limitations of choropleth displays

<!-- # Cancer atlases -->
<!-- Displaying of cancer data on choropleth maps -->
A cancer atlas is a map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, and geographic correlations [@MACM]. The data collection methods across regions and the administrative control within regions lends itself to choropleth visualization. Cancer maps and atlases date back to Haviland's maps in 1875, with more modern atlases directly evolving from early work in US cancer atlases, appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power and the availability of geographic information systems software [@SE].

@@

# Traditional approaches for cancer map displays {#ch:choropleth-maps}
<!-- 381 words -->
<!-- what is a choropleth map -->
A choropleth map displays the geographical distribution of data over a set of spatial units by shading areas of a map [@EI], [@BCM]. Faithful rendering of the geography, when combined with an appropriate color scheme, can reveal spatial patterns among data values. Identifying and explaining spatial structures, patterns, and processes involve considering the individuals and organizing them into representable units of communities [@SAMGIS]. Early versions of choropleth maps used symbols or patterns instead of color. Choropleth maps can be used for displaying disease data [@DMAHP], including cancer data [@CPISACA]. In epidemiology, choropleth maps are often used as a tool to study the spatial distribution of cancer incidence and mortality.

@@

There is large variation in the resolution of the maps. Figure 1<!--\ref{fig:choropleth-grid-create}-->b shows global information at a national level. The United States Cancer Statistics [@USInteractive] shows data aggregated at the state level (n = 51). The Environment and Health Atlas of England and Wales [@EnvEnglandWales2] (Figure 1<!--\ref{fig:choropleth-grid-create}-->a) shows the relative risk for women developing lung cancer  at a neighborhood (small-area) scale (n = 8850). The Asutralian Cancer Atlas (Figure 1<!--\ref{fig:choropleth-grid-create}-->d) shows the relative incidence ratio of lung cancer in males for each Statistical Area at Level 2 [@abs2016] (n = 2292). The Atlas of Cancer in Queensland (Figure 1<!--\ref{fig:choropleth-grid-create}-->c) shows a subset of the Australian Statistical Areas at Level 2 [@abs2016] located in the state of Queensland within Australia [@QLDcancerAtlas] (n = 2292).

@@

Maps can also be used to focus on demographic strata, such as age and sex. Some of the digital atlases surveyed allow subsets such as males, females, or those aged over 65, to be selected for display. Similarly, socioeconomic indicators, such as unemployment rates, poverty rates, remoteness, and education levels, can be used to filter data, in order to communicate how cancer prevalence varies for different members of society. Few atlases provide this level of detail.

Introducing population and demographic information helps to interpret the rates in areas effectively, but there will still be uncertainty around the rates. To address this, a cancer atlas often communicates uncertainty about the value of a statistic. There are several potential sources of uncertainty: sampling error, errors arising from the disease reporting process (or data collection), and errors arising from the statistical modelling or simulation process. The most common measures used to present uncertainty are credible or confidence intervals (CIs). Displaying the uncertainty associated with reported statistics is a vital feature of a cancer map, but it is difficult to display effectively. The map focuses on displaying the statistic and lacks additional space to represent the uncertainty. Providing an adjacent map or overlaying maps with symbols [@VSSDCUC] are two common solutions.

>>>>

<!-- # Cancer atlases -->
<!-- Displaying of cancer data on choropleth maps -->
A cancer atlas is a map, or collection of maps, commoly representing cancer incidence or mortality patterns across a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, and geographic correlations [@MACM]. The data collection methods across regions and the administrative control within regions lends itself to choropleth visualization. Cancer maps and atlases date back to Haviland's maps in 1875, with more modern atlases directly evolving from early work in US cancer atlases, appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power and the availability of geographic information systems software [@SE].

@@

# Visualisation approaches

# Traditional approaches for cancer map displays {#ch:choropleth-maps}

<!-- 381 words -->
<!-- what is a choropleth map -->
A choropleth map displays the geographical distribution of data over a set of spatial units by shading areas of a map [@EI], [@BCM]. Faithful rendering of the geography, when combined with an appropriate color scheme, can reveal spatial patterns among data values. Identifying and explaining spatial structures, patterns, and processes involve considering the individuals and organizing them into representable units of communities [@SAMGIS]. Early versions of choropleth maps used symbols or patterns instead of color. Choropleth maps can be used for displaying disease data [@DMAHP], including cancer data [@CPISACA]. In epidemiology, choropleth maps are often used as a tool to study the spatial distribution of cancer incidence and mortality.

@@

There is large variation in the resolution of the maps. Figure 1<!--\ref{fig:choropleth-grid-create}-->b shows global information at a national level. The United States Cancer Statistics [@USInteractive] shows data aggregated at the state level (n = 51). The Environment and Health Atlas of England and Wales [@EnvEnglandWales2] (Figure 1<!--\ref{fig:choropleth-grid-create}-->a) shows the relative risk for women developing lung cancer  at a neighborhood (small-area) scale (n = 8850). The Australian Cancer Atlas (Figure 1<!--\ref{fig:choropleth-grid-create}-->d) shows the relative incidence ratio of lung cancer in males for each Statistical Area at Level 2 [@abs2016] (n = 2292). The Atlas of Cancer in Queensland (Figure 1<!--\ref{fig:choropleth-grid-create}-->c) shows a subset of the Australian Statistical Local Areas (SLAs) located in the state of Queensland within Australia [@QLDcancerAtlas] (n = 478).

@@

Maps can also be used to focus on demographic strata, such as age and sex. Some of the digital atlases surveyed allow subsets to be selected for display, The Australian Cancer Atlas [@ACA] allows filtering for the selection of statistics regarding males and females specifically. Similarly, remoteness [@ACA], socioeconomic indicators, such as unemployment rates, poverty rates, and education levels, can be used to filter data, in order to communicate how cancer prevalence varies for different population subgroups. Few atlases provide this level of detail. XXX

Introducing population and demographic information helps to interpret the rates in areas effectively, but there will still be uncertainty around the rates. To address this, a cancer atlas often communicates uncertainty about the value of a statistic. There are several potential sources of uncertainty: sampling error, errors arising from the disease reporting or data collection processes, and uncertainty arising from the statistical modelling or simulation process. The most common measures used to present uncertainty are credible intervals or confidence intervals. Displaying the uncertainty associated with reported statistics is a vital feature of a cancer map, but it is difficult to display effectively. Providing an adjacent map or overlaying maps with symbols [@VSSDCUC] are two common solutions.

### 2020-05-15 submission2/submission2.Rmd
Edited according to reviewers suggestions up to Limitations of choropleth displays

**Reviewer suggestion:**
**Suggest also noting development of statistical methods and software**

@@

Several suggestions were made regarding Table 2: Common measures for reporting cancer information.
The additional information requested is communicated more effectively using a paragraph style format.

>>>>

The following reviewer suggestions relate to the introduction:

**Reviewer suggestion:**
**Given the ethical and privacy considerations for small area data, it is not a straightforward process to obtain and report area-specific counts.**

We agree that it is not easy to obtain and report area-specific counts.
The statement has been removed.

**Reviewer suggestion:**
**Suggest moving paragraph to proposed choropleth maps section**

This paragraph has been moved the choropleth section under visualisation approches.

**Reviewer suggestion:**
**Can you include mention of how this is different to a thematic map?**

We have included in the choropleth section:
Choropleth maps are a type of thematic map that show polygons for each of the groups of data points representing the geographic units, where each polygon is shaded with a color according to the area-specific values of the statistic being conveyed.

**Reviewer suggestion:**
**Suggest also noting development of statistical methods and software**
The paragraph that began the *Cancer Atlases* section has been moved to the introduction. This paragraph describes cancer atlases, and the developments oof statistical methods and software.

@@

Several suggestions were made regarding Table 2: Common measures for reporting cancer information.
Additional columns have been included.

XXX

**Reviewer suggestion:**
**This is the wrong reference.**
The incorrect reference to the use of Statistical Areas at Level 2 for the Atlas of Cancer in Queensland has been corrected to reflect that the atlas used Statistical Local Areas.

**Reviewer suggestion:**
The following paragraph was included at the suggestion of the reviewer:

As data becomes more sparse, either by increasing geographic resolution or stratifying by age/sex, obtaining reliable estimates becomes more difficult. Many atlases analyzing smaller areas use statistical modelling to produce estimates, although some continue to use simple calculations and suppress regions where estimates are unstable (e.g. (mention some)). XXX

**Reviewer suggestion:**
**These are very different terms, so ideally should not have the same abbreviation. **
The terms credible intervals and confidence intervals are no longer abbreviated.

**Reviewer suggestion:**
**Which map?**
The following sentence has been removed: The map focuses on displaying the statistic and lacks additional space to represent the uncertainty. 
As the Australian Cancer Atlas example presented in Figure 1 does include uncertainty using transparency.
